Near instantaneous response of mesophyll conductance to light availability allows shade leaves to lie in wait

### 2015-08-03 manuscript/manuscript_wtc3.Rmd
changes to methods

# Introduction
The need to understand and predict carbon (C) uptake in forest ecosystems is crucially important for assessing the impacts of a changing climate. Specifically, this involves accurate upscaling of leaf level measurements to predict whole canopy function. Due to the costs and limitations of efficient light harvesting plants cannot expose all leaves to full sun [@niinemets2010review], making simple upscaling problematic. Incident PPFD declines exponentially with increasing leaf area index, creating a steep light gradient from the canopy top to bottom [@monsi2005factor], and photosynthesis increases with irradiance from shade to sun until biochemical limitation occurs [@evans1995carbon]. Consequently, leaves modify structural and functional properties to efficiently intercept variable light and enhance photosynthetic capacity [@vogelman1996leaves; @valladares2004photosynthetic]. To estimate whole canopy C gain it is thus necessary to account for the non-linear response of photosynthesis to light by distinguishing between shaded and sunlit leaves [@de1997simple; @linderson2012up].

@@

\
Additionally, assessing shade leaf behaviour is made difficult with accounting of short term light fluctuations within a canopy, via sunflecks. How shade leaves utilize sunflecks for short term C gain depends on the combined response time of g~s~ and g~m~ and the underlying photosynthetic biochemistry acclimated to a low light environment [@pearcy1990sunflecks]. For example, at short timescales the utilization of sunflecks is limited by the slow response of stomatal opening, effectively limiting the maximum assimilation rate that can be achieved [@way2012sunflecks]. Mesophyll conductance has been shown to respond to environmental factors across both long and rapid time scales, possibly faster than g~s~ [@flexas2008mesophyll]. However, anatomical parameters which regulate g~m~ with changing irradiance such as chloroplast surface area [@terashima2006irradiance] and mesophyll thickness [@boardman1977comparative] are not likely adaptable during short light fluctuations. The physiological behaviour of shade leaves to maximize C gain must be assessed as both a degree of acclimation to local irradiance and as a potential response to transitory light availability.

\
Climate warming may also potentially affect the physiological behaviour of leaves within a canopy. This is because leaves can be exposed to different heat, water and high light stresses as temperature and vapour pressure deficit (VPD) scale positively with canopy light availability [@baldocchi2002environment; @valladares2004photosynthetic; @niinemets2007photosynthesis]. How these stresses affect the diffusion of CO~2~, through either g~s~ or g~m~ will have implications for upscaling A for sun and shade leaves. Additionally, light saturated rates of A are limited by the maximum rate of Rubisco carboxylation (Vc~max~) or the maximum rate of photosynthetic electron transport (J~max~) across a range of temperatures, yet their temperature dependencies are not the same [@farquhar1980biochemical; @medlyn2002temperature]. How these parameters are differentially affected by warming may impact constraints of N distribution and leaf photosynthetic capacity across light gradients. The impacts of warming on plant physiological processes are obviously vast, yet differentiating their impacts on leaf physiology within a canopy will be essential for evaluating whole tree responses to a changing climate.

@@

\
1.	If whole tree canopies are optimized for C gain, then leaf N, K and photosynthetic capacity were predicted to be higher in full sun leaves.
2.	Under comparable leaf VPD, g~s~ and A should exhibit theoretical optimal behaviour across both sun and shade leaves, while g~m~ was predicted to scale positively with photosynthetic capacity.
3.	As shade leaves are constrained by their underlying biochemistry and slow physiological responses, increases in A following sunfleck simulations were not expected to reach levels of full sun leaves.
4.	The effects of climate warming was predicted to be greater in sun than shade leaves, seen as increases in stomatal closure and reductions in leaf C gain during summer months.

@@

## Whole tree chamber experimental design
Twelve *Eucalyptus tereticornis* seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, NSW, Australia. Each chamber has a maximum height of 10 m and seedlings were grown for 15 months.  A detailed description of the WTC operation and design is available in [@barton2010whole]. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced a constant elevated air temperature treatment of +3&deg;C (ET, Figure `r airvarsfig`). The CO~2~ concentration inside the chamber was set to match outside air, tracking diurnal changes. Due to the temperature treatment air humidity was not explicitly controlled. Trees were watered weekly with 70 L from March 2013 to November 2013.  From December 2013 to final harvest trees were watered fortnightly with the mean monthly (100 yr) rainfall amount.  In February 2013 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Due to a limited range of data for the drought treatment only well-watered trees are reported here.

\
Before seedlings were planted into each chamber they were maintained under well watered conditions in 35 L pots and kept inside each chamber. This allowed for seedlings to gain sufficient biomass before planting while also allowing them to acclimate to chamber temperature treatments. Seedlings were planted into each chamber after mean seedling height reached 100 cm.  After 2 months, floors were installed 45 cm from the soil surface.  This enabled chamber fluxes of CO~2~ and H~2~O from the whole tree canopy to be monitored. Gas exchange measurements were initiated in October 2013 when trees had both ample height growth and canopy development for realistic canopy light gradients to be measured.  At this point, trees under AT treatment had a mean diameter of 28.2&pm;1.1 mm, height of 348&pm;15.1 cm and an estimated leaf area of 3.9&pm;0.1 m^2^. For ET treatments trees had a mean diameter of 34.1&pm;2.1 mm, height of 418.3&pm;23.1 cm and an estimated leaf area of 6.2&pm;0.2 m^2^.

##Leaf gas exchange, online carbon isotope discrimination and mesophyll conductance
Leaf gas exchange measurements were performed monthly through to the final harvest.  Measurements were taken on a representative sun and shade leaf for each tree during each measurement campaign.  The newest fully expanded leaf from the stem apex was chosen for gas exchange measurements and sun leaves were measured in the upper third of the canopy. In order to minimize any confounding effects of leaf age, shade leaves were measured first in the lower canopy then gradually higher during each campaign.  As shade leaves most likely developed slower this assured that older leaves in the lower canopy were avoided. The nearest leaf on each branch was sampled for measurement of predawn leaf water potential. All leaves were selected and flagged 24 hours prior to initiation of measurements.

\
Prior to gas exchange measurements photosynthetic photon flux density (PPFD) was recorded at the individual leaf level and at the canopy position for each selected leaf.  A hand-held photosynthetically available radiation meter was used to record leaf level PPFD to ensure that chosen leaves were positioned in the desired light environment, either full sun or full shade. A ceptometer (AccuPAR LP-80, Decagon Devices, Pullman, USA) was then used to measure 1 m integrated PPFD at the canopy height of each chosen leaf type. Five ceptometer readings were recorded within the canopy at random locations of the height of each selected leaf.  The mean of these readings was assumed to represent the local light environment of full sun and shade leaves for each tree. All measurements of PPFD and gas exchange were performed on full sun days between 10:00-14:30 h.

@@

# Discussion
Here we show that dynamic physiological responses in shade leaves to altered light environments has important implications for upscaling leaf level measurements to the canopy. Tree canopies may maintain the ability to upregulate A beyond the constraints of their local light environment, even at the cost of valuable resources. Empirical data accounting for the total conductance of CO2 (both g~s~ and g~m~) reveal that these responses can lead to leaf C gain equivalent to sun leaves and may happen quicker than previously thought. Thus, failure to integrate the effects of g~m~ into process based tree growth models may underestimate the ability of canopies to alter their photosynthetic capacity when environmental conditions change. Additionally, with measurements recorded across a large natural range of temperatures we were unable to detect any effects of a +3 &deg;C warming treatment on leaf physiology.

## sub heading
The allocation of N constrains photosynthesis rates and is thus a key trait in determining the relative contribution of individual leaves to canopy C gain.  Decreasing light availability should decrease the investment into photosynthetic enzyme within a canopy (Mooney and Gulmon 1979).  As a result, acclimation of photosynthetic capacity to irradiance is typically reflected in the key photosynthetic biochemical parameters Vcmax and Jmax [@farquhar1980biochemical]. Our data agree with these conventional conclusions as the distribution of N, both biochemical parameters and A were all reduced in shade leaves.

\
Photosynthesis in trees is also limited by the ability to supply water to the upper canopy. Using a two-leaf model Peltoniemi et al. [-@peltoniemi2012co] theorizes that optimal N distribution will be proportional to light distribution only if leaf K is also optimally distributed. In this study, variation in leaf N distribution and A rates were not associated with subsequent changes in leaf K between sun and shade leaves. This is meaningful as foliage tends to operate at greater water limitations in the upper canopy, despite the ability of sun leaves to develop more effective vascular systems [@sellin2008effects;@niinemets2012optimization]. As both sun and shade leaves were acclimated to their local light environment this sub-optimal distribution of water supply has the potential to diminish the C return for the whole canopy.

@@

\
The pattern of wasteful water use in shade leaves is important as we hypothesized that gs and A would be proportional across sun and shade leaves.  In sun leaves variation in A and g~s~ were strongly correlated, exhibiting behaviour agreeing with optimal stomatal theory. However, lower rates of A in shade leaves were not coupled with decreases in g~s~ which resulted in the observed decreases in WUE. This is significant as optimal stomatal behaviour has been reported across a wide range of ecosystems and plant functional types; however, empirical data is often collected only on sun leaves [see @prentice2014balancing; @lin2015optimal].  As a result, the often used economic framework of balancing costs of using water versus N allocation to predict photosynthesis [@wright2003least] may break down when considering all leaves within a tree canopy.

\
This behaviour may be explained by the response of shade leaves to increases in light availability. Contrary to hypothesis, shade leaves were able to increase A to rates equivalent or even outperforming sun leaves under identical light intensity. It is possible that sustaining higher g~s~, at the cost of larger transpiration losses, is a strategy to increase the utilization efficiency of sunflecks. These results should not be entirely unexpected as shade leaves experience interchanging periods of sun and shade, despite canopy light extinction [@pearcy1990sunflecks]. Additionally, empirical studies that assess photosynthetic responses to sunflecks generally focus on seedlings [@kuppers1993leaf;@leakey2002relative] and understory plants [@chazdon1991importance; @brantley2009contribution]. Our findings highlight a critical need for empirical measurements of shade leaves under dynamic light environments in order to accurately scale C gain from leaf to canopy.

\
Increasing the drawdown of CO~2~ into the intercellular air spaces via g~s~, however, does not necessarily infer concomitant increases in A. This is because CO~2~ supply to the site of carboxylation, via g~m~, may be anatomically constrained. In this study A and g~m~ scaled positively across leaf types and increased proportionately with increased light intensity in shade leaves.  Research has suggested that aquaporins can facilitate increases in the CO~2~ permeability of the cell membranes resulting in rapid modulation of mesophyll conductance [@hanba2004overexpression; @heinen2009role; @li2014plant]. This provides a potential explanation for our observed rapid increases in g~m~ but is untested in leaves of woody species. Our findings support a growing wealth of evidence that g~m~ is highly variable and can respond to environmental variables [@flexas2008mesophyll]. Here we provide empirical data showing g~m~ not only varies within a canopy but the up-regulation of g~m~ plays a critical role in the photosynthetic response of shade leaves to sunflecks.

\
Similar to Laisk [-@laisk2005adjustment] we show evidence that optimal acclimation, in this case physiological adjustment of the drawdown of CO~2~ in shade leaves, is possibly directed towards occasional maximums of light availability over adaptations towards a low light environment.  If shade leaves ‘lie in wait’ for sunflecks then perhaps we should consider an alternate leaf economic strategy to maximise C gain, beyond conventional trade-offs associated with canopy resource distribution.  If the valuation of sunflecks as a C resource is large enough then costs of sub-optimal stomatal behaviour in shade leaves could be offset over the leaf lifespan.  The potential benefits of C gain in leaves constrained by low light to levels equivalent to sun leaves may be large enough to tolerate excess losses from transpiration.  However, accounting for the heterogeneous nature of light within a canopy remains a current challenge for empirical and modelling studies. Thus, models which predict leaf photosynthesis from N distribution within a canopy will be flawed unless inclusion of canopy light extinction and the integrations of sunflecks on shade leaves are included [@de1997simple].

\
Furthermore, our findings have important implications for process based models that predict canopy C gain from rates of leaf photosynthesis. Although the distributions of resources, such as N and water, define leaf photosynthetic capacity it is the physiological behaviour of individual leaves which actually determine C gain. The observed physiological behaviour leaves in the shade, at both low and high irradiance, suggests that current theories of leaf optimal behaviour are incomplete.  Furthermore, gm must now be considered as a dynamic process in modelling whole canopy C gain.  Additional empirical data are needed to determine both the mechanisms responsible for and the capacity of gm to rapidly increase CO2 drawdown. To improve our ability to predict whole canopy C gain future research should prioritize the incorporation of both sun and shade leaf physiology, which may be optimized differently.

>>>>

# Introduction
The need to understand and predict carbon (C) uptake in forest ecosystems is crucially important for assessing the impacts of environmental change. Specifically, this involves accurate upscaling of leaf level measurements to predict whole canopy function. Due to the costs and limitations of efficient light harvesting plants cannot expose all leaves to full sun [@niinemets2010review], making simple upscaling problematic. Incident PPFD declines exponentially with increasing leaf area index, creating a steep light gradient from the canopy top to bottom [@monsi2005factor], and photosynthesis (A) increases with irradiance from shade to sun until biochemical limitation occurs [@evans1995carbon]. Consequently, leaves modify structural and functional properties to efficiently intercept variable light and enhance photosynthetic capacity [@vogelman1996leaves; @valladares2004photosynthetic]. To estimate whole canopy C gain it is thus necessary to account for the non-linear response of A to light by distinguishing between shaded and sunlit leaves [@de1997simple; @linderson2012up].

@@

\
Additionally, assessing shade leaf behaviour is made difficult with accounting of short term light fluctuations within a canopy, via sunflecks. How shade leaves utilize sunflecks for short term C gain depends on the combined response time of g~s~ and g~m~ and the underlying photosynthetic biochemistry acclimated to a low light environment [@pearcy1990sunflecks]. For example, at short timescales the utilization of sunflecks is limited by the slow response of stomatal opening, effectively limiting the maximum assimilation rate that can be achieved [@way2012sunflecks]. Mesophyll conductance has been shown to respond to environmental factors across both long and rapid time scales, possibly faster than g~s~ [@flexas2008mesophyll]. However, anatomical parameters which regulate g~m~ with changing irradiance such as chloroplast surface area [@terashima2006irradiance] and mesophyll thickness [@boardman1977comparative] are unlikely to adapt during short light fluctuations. The physiological behaviour of shade leaves to maximize C gain must be assessed as both a degree of acclimation to local irradiance and as a potential response to transitory light availability.

\
Climate warming may also potentially affect the physiological behaviour of leaves within a canopy. This is because leaves can be exposed to different heat, water and high light stresses as temperature and vapour pressure deficit (VPD) scale positively with canopy light availability [@baldocchi2002environment; @valladares2004photosynthetic; @niinemets2007photosynthesis]. How these stresses affect the diffusion of CO~2~, through either g~s~ or g~m~ will have implications for upscaling A for sun and shade leaves. Additionally, light saturated rates of A are limited by the maximum rate of Rubisco carboxylation (Vc~max~) or the maximum rate of photosynthetic electron transport (J~max~) across a range of temperatures, yet their temperature dependencies are not the same [@farquhar1980biochemical; @medlyn2002temperature]. How these parameters are differentially affected by warming may impact constraints of N distribution and leaf photosynthetic capacity across light gradients. The impacts of warming on plant physiological processes are obviously vast, yet differentiating their impacts on leaf physiology within a canopy will be essential to evaluate whole tree responses to a changing climate.

@@

\
1.	If whole tree canopies are optimized for C gain, then leaf N, K and photosynthetic capacity were predicted to be higher in full sun leaves.

\
2.	Under comparable leaf VPD, g~s~ and A should be proportional across both sun and shade leaves, while g~m~ was predicted to scale positively with photosynthetic capacity.

\
3.	As shade leaves are constrained by their underlying biochemistry and slow physiological responses, increases in A following sunfleck simulations were not expected to reach levels of full sun leaves.

\
4.	The effects of climate warming was predicted to be greater in sun than shade leaves, seen as increases in stomatal closure and reductions in leaf C gain during summer months.

@@

## Whole tree chamber experimental design
Twelve *Eucalyptus tereticornis* seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, NSW, Australia. Each chamber has a maximum height of 10 m and seedlings were grown for 15 months.  A detailed description of the WTC operation and design is available in [@barton2010whole]. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced a constant elevated air temperature treatment of +3&deg;C (ET, Figure `r airvarsfig`). Due to the temperature treatment air humidity was not explicitly controlled. The CO~2~ concentration inside the chamber was set to match outside air, tracking diurnal changes.  Trees were watered weekly with 70 L from March 2013 to November 2013.  From December 2013 to final harvest trees were watered fortnightly with the mean monthly (100 yr) rainfall amount.  In February 2013 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Due to a limited range of data for the drought treatment only well-watered trees are reported here.

\
Before seedlings were planted into each chamber they were maintained under well watered conditions in 35 L pots and kept inside each chamber. This allowed for seedlings to gain sufficient biomass before planting while also allowing them to acclimate to chamber temperature treatments. Seedlings were planted into each chamber after mean seedling height reached 100 cm.  The top soils at this site, used in both pots and chambers, are an alluvial formation of low-fertility sandy loam soils with low organic matter and low water holding capacity. After 2 months, floors were installed 45 cm from the soil surface.  This enabled chamber fluxes of CO~2~ and H~2~O from the whole tree canopy to be monitored. Gas exchange measurements were initiated in October 2013 when trees had both ample height growth and canopy development for realistic canopy light gradients to be measured.  At this point, trees under AT treatment had a mean diameter of 28.2&pm;1.1 mm, height of 348&pm;15.1 cm and an estimated leaf area of 3.9&pm;0.1 m^2^. For ET treatments trees had a mean diameter of 34.1&pm;2.1 mm, height of 418.3&pm;23.1 cm and an estimated leaf area of 6.2&pm;0.2 m^2^.

##Leaf gas exchange, online carbon isotope discrimination and mesophyll conductance
Leaf gas exchange measurements were performed monthly through to the final harvest.  Measurements were taken on a representative sun and shade leaf for each tree during each measurement campaign.  The newest fully expanded leaf from the stem apex was chosen for gas exchange measurements and sun leaves were measured in the upper third of the canopy. In order to minimize any confounding effects of leaf age, shade leaves were measured first in the lower canopy then gradually higher during each campaign.  As shade leaves most likely developed slower this assured that older leaves in the lower canopy were avoided. The nearest leaf on each branch was sampled for measurement of predawn leaf water potential.
\
Prior to gas exchange measurements photosynthetic photon flux density (PPFD) was recorded at the individual leaf level and at the canopy position for each selected leaf.  A hand-held quantum sensor () was used to record leaf level PPFD to ensure that chosen leaves were positioned in the desired light environment, either full sun or full shade. A ceptometer (AccuPAR LP-80, Decagon Devices, Pullman, USA) was then used to measure 1 m integrated PPFD at the canopy height of each chosen leaf type. Five ceptometer readings were recorded within the canopy at random locations of the height of each selected leaf.  The mean of these readings was assumed to represent the local light environment of full sun and shade leaves for each tree. All measurements of PPFD and gas exchange were performed on full sun days between 10:00-14:30 h.

@@

# Discussion
Here we show that dynamic physiological responses in shade leaves to altered light environments has important implications for upscaling leaf level measurements to the canopy. Tree canopies may maintain the ability to upregulate A beyond the constraints of their local light environment, even at the cost of valuable resources. Empirical data accounting for the total conductance of CO~2~ reveal that shifts in g~s~ and g~m~ in shade leaves can lead to leaf C gain equivalent to sun leaves and may happen quicker than previously thought. Thus, failure to integrate the effects of g~m~ into process based tree growth models may underestimate the ability of canopies to alter their photosynthetic capacity when environmental conditions change. Additionally, with measurements recorded across a large natural range of temperatures we were unable to detect any effects of a +3 &deg;C warming treatment on leaf physiology.

## sub heading
The allocation of N constrains photosynthesis rates and is thus a key trait in determining the relative contribution of individual leaves to canopy C gain.  Decreasing light availability should decrease the investment into photosynthetic enzyme within a canopy (Mooney and Gulmon 1979).  As a result, acclimation of photosynthetic capacity to irradiance is typically reflected in the key photosynthetic biochemical parameters Vc~max~ and J~max~ [@farquhar1980biochemical]. Our data agree with these conventional conclusions as the distribution of N, both biochemical parameters and A were all reduced in shade leaves.

\
Photosynthesis in trees is also limited by the ability to supply water to the upper canopy. Using a two-leaf model Peltoniemi et al. [-@peltoniemi2012co] theorizes that optimal N distribution will be proportional to light distribution only if leaf K is also optimally distributed. In this study, variation in leaf N distribution and A rates were not associated with subsequent changes in leaf K between sun and shade leaves. This is especially meaningful in trees as foliage generally operates at greater water limitations in the upper canopy. As both sun and shade leaves were acclimated to their local light environment this sub-optimal distribution of water supply has the potential to diminish the C return for the whole canopy.

@@

\
The pattern of wasteful water use in shade leaves is important as we hypothesized that g~s~ and A would be proportional across sun and shade leaves.  In sun leaves variation in A and g~s~ were strongly correlated, exhibiting behaviour agreeing with optimal stomatal theory. However, lower rates of A in shade leaves were not coupled with decreases in g~s~, explaining the observed decreases in WUE. This is significant as optimal stomatal behaviour has been reported across a wide range of ecosystems and plant functional types; however, empirical data is often collected only on sun leaves [see @prentice2014balancing; @lin2015optimal].  As a result, the often used economic framework of balancing costs of using water versus N allocation to predict A [@wright2003least] may break down when considering all leaves within a tree canopy.

\
This behaviour may be explained by the response of shade leaves to increases in light availability. Contrary to hypothesis, shade leaves were able to increase A to rates equivalent or even outperforming sun leaves under identical light intensity. It is possible that sustaining higher g~s~, at the cost of larger transpiration losses, is a strategy to increase the utilization efficiency of sunflecks. These results should not be entirely unexpected as shade leaves experience interchanging periods of sun and shade, despite canopy light extinction [@pearcy1990sunflecks]. Unfortunately, empirical studies that assess photosynthetic responses to sunflecks generally focus on seedlings [@kuppers1993leaf;@leakey2002relative] and understory plants [@chazdon1991importance; @brantley2009contribution]. Our findings highlight a critical need for empirical measurements of shade leaves under dynamic light environments in order to accurately scale C gain from leaf to canopy.

\
Increasing the drawdown of CO~2~ into the intercellular air spaces via g~s~, however, does not necessarily infer concomitant increases in A. This is because CO~2~ supply to the site of carboxylation, via g~m~, may be anatomically constrained. In this study A and g~m~ scaled positively across leaf types and increased proportionately with increased light intensity in shade leaves.  Research has suggested that aquaporins can facilitate increases in the CO~2~ permeability of the cell membranes resulting in rapid modulation of mesophyll conductance [@hanba2004overexpression; @heinen2009role; @li2014plant]. This provides a potential explanation for the observed rapid increases in g~m~ but is laregly untested in leaves of tree species. Our findings support a growing wealth of evidence that g~m~ is highly variable and can respond to environmental variables [@flexas2008mesophyll]. Here we provide empirical data showing g~m~ not only varies within a canopy but the up-regulation of g~m~ plays a critical role in the photosynthetic response of shade leaves to sunflecks.

\
Similar to Laisk [-@laisk2005adjustment] we provide evidence that optimal acclimation, in this case physiological adjustment of the drawdown of CO~2~ in shade leaves, is possibly directed towards occasional maximums of light availability over adaptations towards a low light environment.  If shade leaves "lie in wait" for sunflecks then perhaps we should consider an alternate leaf economic strategy to maximise C gain, beyond conventional trade-offs associated with canopy resource distribution.  If the valuation of sunflecks as a C resource is large enough then costs of sub-optimal stomatal behaviour could be offset over the leaf lifespan. For example, the potential benefits of C gain in leaves constrained by low light to levels equivalent to sun leaves may be large enough to tolerate excess losses from transpiration.  However, accounting for the heterogeneous nature of light within a canopy remains a current challenge for empirical and modelling studies. Thus, models which predict leaf photosynthesis from N distribution within a canopy will be flawed unless inclusion of canopy light extinction and the integrations of sunflecks on shade leaves are included [@de1997simple].

\
Furthermore, our findings have important implications for process based models that predict canopy C gain from rates of leaf photosynthesis. Although the distributions of resources, such as N and water, define leaf photosynthetic capacity it is the physiological behaviour of individual leaves which actually determine C gain. The observed physiological behaviour leaves in the shade, at both low and high irradiance, suggests that current theories of leaf optimal behaviour are incomplete.  Furthermore, g~m~ must now be considered as a dynamic process in modelling whole canopy C gain.  Additional empirical data are needed to determine both the mechanisms and the capacity of g~m~ to rapidly increase CO~2~ drawdown. To improve our ability to predict whole canopy C gain future research should prioritize the incorporation of both sun and shade leaf physiology, which may be optimized differently.
